---
title: "Bayesian SIR Model in NIMBLE"
subtitle: "STAT 244NF: Infectious Disease Modeling"
author: "YOUR NAME HERE"
output: pdf_document
---

# Recap -- Bayesian SIR Model

In the previous class, we discussed broadly the general Bayesian framework and connected it specifically to a stochastic SIR model. In our notes, we considered an SIR model with two parameters, $\boldsymbol{\theta}=(\theta_1, \theta_2)$, which were the probability of an $S \rightarrow I$ transition and the probability of an $I \rightarrow R$ transition, respectively. In that setup, we kept these parameters constant over time. In this lab, we will consider a slightly more useful formulation of the SIR model. It is similar to formulations we have considered in this case, with one modification of the $S \rightarrow I$ transition probability and some small notation modifications. 

# Model Setup

## Differencing Equations

$$S_{t}=S_{t-1}-I_t^{(SI)}$$
$$I_{t}=I_{t-1}+I_t^{(SI)}-R_t^{(IR)}$$
$$R_{t}=R_{t-1}+R_t^{(IR)}$$

## Data Model $f(\mathbf{X}|\boldsymbol{\theta})$

For all $t \in 1,...,\tau$,

$$I_t^{(SI)}|\beta\sim \text{Binomial}\left(S_{t-1}, 1-e^{-\beta\times I_{t-1}/N} \right)$$
$$R_t^{(IR)}|\gamma\sim \text{Binomial}\left(I_{t-1}, 1-e^{-\gamma} \right)$$

## Prior Model $\pi(\boldsymbol{\theta})$

$$\beta \sim \text{Gamma}\left(0.01, 0.01 \right)$$
$$\gamma \sim \text{Gamma}\left(a, b \right)$$

# Writing the SIR model

```{r}
library(nimble)
```


```{r}
SIR_code <-  nimbleCode({
    
    S[1] <- N - I0 - R0
    I[1] <- I0
    R[1] <- R0
    
    probIR <- 1 - exp(-gamma)
    
    ### loop over time
    for(t in 1:tau) {
        
      ## indexing is a little different from what we have used, but it is the same idea
      ## we do not have to create storage here
        probSI[t] <- 1 - exp(- beta * I[t] / N)
        
        I_SI[t] ~ dbin(probSI[t], S[t]) ## follows a Binomial distribution
        R_IR[t] ~ dbin(probIR, I[t]) ## follows a Binomial distribution
        
        # update S, I, R
        S[t + 1] <- S[t] - I_SI[t]
        I[t + 1] <- I[t] + I_SI[t] - R_IR[t]
        R[t + 1] <- R[t] + R_IR[t] 
        
    }
    
    # priors
    beta ~ dgamma(0.1, 0.1) ## transmission rate
    gamma ~ dgamma(aa, bb) ## recovery rate
    
})
```


# Simulating epidemics

Here we specify the population size $N = 10,000$, 5 initially infectious individuals,
and simulate 100 days of the epidemic. 

```{r}
constantsList <- list(N = 10000,
                      I0 = 5,
                      R0 = 0,
                      tau = 100)


sirModel <- nimbleModel(SIR_code, 
                        constants = constantsList)

# exclude data from parent nodes
dataNodes <- c('I_SI', 'R_IR')
dataNodes <- sirModel$expandNodeNames(dataNodes, returnScalarComponents = TRUE)
parentNodes <- sirModel$getParents(dataNodes, stochOnly = TRUE)
parentNodes <- parentNodes[-which(parentNodes %in% dataNodes)]
parentNodes <- sirModel$expandNodeNames(parentNodes, returnScalarComponents = TRUE)
nodesToSim <- sirModel$getDependencies(parentNodes, self = FALSE, downstream = T)



```


We can simulate using various values of $\beta$ and $\gamma$ to specify various
reproductive numbers.

In all simulations the mean infectious period is 5 days.


```{r, fig.width = 12, fig.height=12}
pal <- c('forestgreen', 'red', 'blue')


par(mfrow = c(2,2))

# simulation 1
initsList <- list(beta = 0.8,
                  gamma = 0.2)
sirModel$setInits(initsList)

set.seed(1)
sirModel$simulate(nodesToSim, includeData = TRUE)

plot(sirModel$S, type = 'l', col = pal[1], ylim = c(0, 1.3e4),
     main = paste0('R0 = ', sirModel$beta / sirModel$gamma), lwd = 2)
lines(sirModel$I, col = pal[2], lwd = 2)
lines(sirModel$R, col = pal[3], lwd = 2)
legend('topright', c('S', 'I', 'R'), col = pal, lwd = 2,
       bty = 'n', horiz = T)

# simulation 2
initsList <- list(beta = 0.6,
                  gamma = 0.2)
sirModel$setInits(initsList)

set.seed(1)
sirModel$simulate(nodesToSim, includeData = TRUE)

plot(sirModel$S, type = 'l', col = pal[1], ylim = c(0, 1.3e4),
     main = paste0('R0 = ', sirModel$beta / sirModel$gamma), lwd = 2)
lines(sirModel$I, col = pal[2], lwd = 2)
lines(sirModel$R, col = pal[3], lwd = 2)
legend('topright', c('S', 'I', 'R'), col = pal, lwd = 2,
       bty = 'n', horiz = T)

# simulation 3
initsList <- list(beta = 0.4,
                  gamma = 0.2)
sirModel$setInits(initsList)

set.seed(1)
sirModel$simulate(nodesToSim, includeData = TRUE)

plot(sirModel$S, type = 'l', col = pal[1], ylim = c(0, 1.3e4),
     main = paste0('R0 = ', sirModel$beta / sirModel$gamma), lwd = 2)
lines(sirModel$I, col = pal[2], lwd = 2)
lines(sirModel$R, col = pal[3], lwd = 2)
legend('topright', c('S', 'I', 'R'), col = pal, lwd = 2,
       bty = 'n', horiz = T)

# simulation 4
initsList <- list(beta = 0.2,
                  gamma = 0.2)
sirModel$setInits(initsList)

set.seed(1)
sirModel$simulate(nodesToSim, includeData = TRUE)

plot(sirModel$S, type = 'l', col = pal[1], ylim = c(0, 70),
     main = paste0('R0 = ', sirModel$beta / sirModel$gamma), lwd = 2)
lines(sirModel$I, col = pal[2], lwd = 2)
lines(sirModel$R, col = pal[3], lwd = 2)
legend('topright', c('S', 'I', 'R'), col = pal, lwd = 2,
       bty = 'n', horiz = T)
```

# Epidemics are Stochastic

Here we simulate 100 epidemics from the same parameter values and plot
the observed incidence curve from each simulation.

```{r}
initsList <- list(beta = 0.4,
                  gamma = 0.2)
sirModel$setInits(initsList)

nSim <- 100

set.seed(1)
epiCurve <- matrix(NA, nrow = length(sirModel$I_SI), ncol = nSim)
for (i in 1:nSim) {
    sirModel$simulate(nodesToSim, includeData = TRUE)
    epiCurve[,i] <- sirModel$I_SI
}

plot(epiCurve[,1], type = 'l', col = adjustcolor('black', alpha = 0.3),
     ylim = c(0, 500))
for (i in 2:nSim) {
    lines(epiCurve[,i], col = adjustcolor('black', alpha = 0.3))
}

```

# Model fitting to simulated data

Simulate data, then use it to fit the model.

```{r}
initsList <- list(beta = 0.6,
                  gamma = 0.2)
sirModel$setInits(initsList)

set.seed(1)
sirModel$simulate(nodesToSim, includeData = TRUE)

trueI_SI <- sirModel$I_SI
trueR_IR <- sirModel$R_IR

endTime <- max(which(trueI_SI > 0)) + 10
trueI_SI <- trueI_SI[1:endTime]
trueR_IR <- trueR_IR[1:endTime]

plot(trueI_SI, type = 'l')
lines(trueR_IR, col = 'red')
legend('topright', c('incidence', 'removals'), col = c('black', 'red'), lwd = 1)
```


# Model Specifications

Before fitting the model, we need to determine a reasonable prior for $\gamma$.
The true value corresponds to a mean infectious period of 5 days, so we choose a
prior that puts 90% probability on the mean infectious period between 4 and 6 days
and is centered on 5 days.

```{r}
bb <- 348
aa <- 0.2 * bb
pgamma(1/4, aa, bb) - pgamma(1/6, aa, bb)
curve(dgamma(x, aa, bb))
```


```{r}
dataList <- list(I_SI = trueI_SI,
                 R_IR = trueR_IR)

constantsList <- list(N = 10000,
                      I0 = 5,
                      R0 = 0,
                      tau = length(dataList$I_SI),
                      aa = aa,
                      bb = bb) 

set.seed(2)
initsList <- list(beta = runif(1, 0, 1),
                  gamma = rgamma(1, aa, bb))


sirModelFit <- nimbleModel(SIR_code, 
                           constants = constantsList,
                           data = dataList,
                           inits = initsList)



```

NIMBLE automatically calculates S, I, and R from I_SI and R_IR, so these do 
not need to be inputs to the model

```{r}
with(sirModelFit, cbind(S, I_SI, I, R_IR, R))[1:20,]
```



# Use Default Configurations and Obtain Samples

Plotted with burn-in included here

```{r}
myConfig <- configureMCMC(sirModelFit)
myMCMC <- buildMCMC(myConfig)

system.time({
    compiled <- compileNimble(sirModelFit, myMCMC) 
    samples <- runMCMC(compiled$myMCMC, niter = 50000, setSeed = 3)
})

head(samples)

par(mfrow = c(1,2))
plot(samples[,'beta'], type = 'l')
abline(h = 0.6, col = 'red')
plot(samples[,'gamma'], type = 'l')
abline(h = 0.2, col = 'red')

```


*This lab was adapted from Dr. Caitlin Ward's ISBA tutorial on Epidemic Modelling (https://github.com/ceward18/ISBA-2022-Epidemic-Modelling-Workshop/tree/main). *